---
output: 
  github_document:
    toc: TRUE

---



<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# *README: Below, is readme that provides steps for building a package.  This readme acts as a checklist, and control document as functions used in package building are included. 🚧 ✅ *


# Introducing the {ggwaterfall} package! 

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


The goal of {ggwaterfall} is to make visualizing in-out flows with waterfall charts easier.


To install the dev version use the following:

```
remotes::install_github("EvaMaeRey/ggwaterfall") 
```  
  
## Example

Try in an interactive session:  🦄 🦄 🦄 

```{r}
library(ggwaterfall)

library(tidyverse)
flow_df <- data.frame(event = c(
                     "Sales", 
                     "Refunds",
                     "Payouts", 
                     "Court Losses", 
                     "Court Wins", 
                     "Contracts", 
                     "Fees"),
           change = c(6400, -1100, 
                      -100, -4200, 3800, 
                      1400, -2800)) |> 
  mutate(event = factor(event))

flow_df

flow_df |> 
  ggplot() +
  geom_hline(yintercept = 0) +
  aes(change = change, 
      x = event) + # event in order
  geom_waterfall() + 
  geom_waterfall_label() + 
  scale_y_continuous(expand = expansion(.1)) + 
  scale_fill_manual(values = c("springgreen4", "darkred"))

last_plot() + 
  aes(x = fct_reorder(event, change)) # in order neg to positive

last_plot() + 
  aes(x = fct_reorder(event, abs(change))) 
```



# Part I. Work out functionality  ✅ 

## Step 0. use base ggplot2 to get job done

This is by and large taken from https://vita.had.co.nz/papers/ggplot2-wires.pdf, but I removed 'starting cash' and 'grand total', because they don't seem like-in-kind with the other events.  I starting cash could be added via an argument (and always put in the x = 1 position), and grand total could be calculated (argument grand_total = T).

```{r cars}
library(tidyverse)
data.frame(event = c(
                     "Sales", 
                     "Refunds",
                     "Payouts", 
                     "Court Losses", 
                     "Court Wins", 
                     "Contracts", 
                     "Fees"),
           change = c(6400, -1100, 
                      -100, -4200, 3800, 
                      1400, -2800)) %>% 
  mutate(event = factor(event)) ->
flow_df
  
flow_df %>%   # maybe add factor in order if factor is not defined...
  mutate(x_pos = event %>% as.numeric()) %>% 
  arrange(x_pos) %>% 
  mutate(balance = cumsum(c(0, 
                            change[-nrow(.)]))) %>% 
  mutate(flow = factor(sign(change))) ->
balance_df

ggplot(balance_df) +
          geom_rect(
            aes(xmin = x_pos - 0.45, 
                xmax = x_pos + 0.45, 
                ymin = balance, 
                ymax = balance + change)) +
          geom_text(aes(x = event, 
                        y = pmin(balance, 
                                 balance + change) - 50, 
                        label = balance))
```


## Step 1. & 2. write compute functions and pass to ggproto

```{r StatWaterfall}
#' Title
#'
#' @param data 
#' @param scales 
#' @param width 
#'
#' @return
#' @export
#'
#' @examples
compute_panel_waterfall <- function(data, scales, width = .90){
  
  data %>% 
  mutate(x_scale = x) %>% 
  mutate(x_pos = x %>% as.numeric()) %>% 
  arrange(x_pos) %>% 
  mutate(balance = cumsum(c(0, 
                            change[-nrow(.)]))) %>% 
  mutate(direction = factor(sign(change))) %>% 
  mutate(xmin = x_pos - width/2,
         xmax = x_pos + width/2,
         ymin = balance,
         ymax = balance + change) %>% 
  mutate(x = x_pos) %>% 
  mutate(y = ymax) %>% 
  mutate(gain_loss = ifelse(direction == 1, "gain", "loss"))
  
}


### Step 1.1 Test compute 

# flow_df %>% 
#   rename(x = event) %>% 
#   compute_panel_waterfall() 



## Step 2. Pass compute to ggproto 

StatWaterfall <- ggplot2::ggproto(`_class` = "StatWaterfall", 
                         `_inherit` = ggplot2::Stat,
                         required_aes = c("change", "x"),
                         compute_panel = compute_panel_waterfall,
                         default_aes = ggplot2::aes(label = ggplot2::after_stat(change),
                                           fill = ggplot2::after_stat(gain_loss),
                                           vjust = ggplot2::after_stat((direction == -1) %>%
                                                                as.numeric)))
```


## Step 3. pass to geom_ / stat_

```{r geom_waterfall}
#' Title
#'
#' @param geom 
#' @param mapping 
#' @param data 
#' @param position 
#' @param na.rm 
#' @param show.legend 
#' @param inherit.aes 
#' @param ... 
#'
#' @return
#' @export
#'
#' @examples
stat_waterfall <- function(geom = ggplot2::GeomRect, 
  mapping = NULL,
  data = NULL,
  position = "identity",
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatWaterfall,  # proto object from step 2
    geom = geom,  # inherit other behavior
    data = data,
    mapping = mapping,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}

#' Title
#'
#' @param geom 
#' @param mapping 
#' @param data 
#' @param position 
#' @param na.rm 
#' @param show.legend 
#' @param inherit.aes 
#' @param ... 
#'
#' @return
#' @export
#'
#' @examples
geom_waterfall <- stat_waterfall


#' Title
#'
#' @param mapping 
#' @param data 
#' @param position 
#' @param na.rm 
#' @param show.legend 
#' @param inherit.aes 
#' @param ... 
#'
#' @return
#' @export
#'
#' @examples
geom_waterfall_label <- function(..., lineheight = .8){
  stat_waterfall(geom = "text", 
                 lineheight = lineheight, ...)}
```


## Step 4. Enjoy (test)


```{r}
flow_df %>% 
  ggplot() +
  geom_hline(yintercept = 0) +
  aes(change = change, 
      x = event) + # event in order
  geom_waterfall() + 
  geom_waterfall_label() + 
  scale_y_continuous(expand = expansion(.1)) + 
  scale_fill_manual(values = c("springgreen4", "darkred"))

last_plot() + 
  aes(x = fct_reorder(event, change)) # in order neg to positive

last_plot() + 
  aes(x = fct_reorder(event, abs(change))) # in order magnitude
```  

## A non monetary example... 

We can also think about things that 'fill us up' or don't.   Here's an example about building packages... 

```{r}
tibble::tribble(~activity, ~change,
  "writing functions :-)", 50,
  "trying functions out :-)", 180,
  "writing formal tests :-(", -120
  ) |>
ggplot() + 
  aes(x = activity %>% 
        str_wrap(12) %>% 
        fct_inorder(), change = change) + 
  geom_waterfall(show.legend = F) + 
  geom_waterfall_label(aes(label = ifelse(change < 0, 
                       "tedious", 
                       "delightful")), size = 5) +
  labs(title = "What affect do the develpment activities\nhave on your mood and energy?") + 
  geom_hline(yintercept = 0, linetype = "dashed", alpha = .2) +
  
  theme_minimal(base_size = 15)
```


# Part II. Packaging and documentation  🚧 ✅ 

## Phase 1. Minimal working package

### Created files for package archetecture with `devtools::create(".")` ✅ 

### Moved functions R folder? ✅  

```{r}
knitr::knit_code$get() |> names()
```

Use new {readme2pkg} function to do this from readme...

```{r}
readme2pkg::chunk_to_r(c(
                         "StatWaterfall",
                         "geom_waterfall")
                       )
```



### Added roxygen skeleton? ✅ 

Use a roxygen skeleton for auto documentation and making sure proposed functions are *exported*.

### Managed dependencies ? ✅ 

Package dependencies managed, i.e. `depend::function()` in proposed functions and declared in the DESCRIPTION

```{r}
usethis::use_package("ggplot2")
```

### Chosen a license? ✅ 

```{r}
usethis::use_mit_license()
```

### Run `devtools::check()` and addressed errors? ✅ 

```{r, results='hide', error=T}
devtools::check(pkg = ".")
```

### Build package 🚧 

```{r}
devtools::build()
```

You need to do this before Part 0 in this document will work.

### Make aspirational part of readme real. 🚧 

At this point, you could change *eval* chunk options to TRUE. You can remove  the 🦄 emoji and perhaps replace it with construction site if you are still uncertain of the API, and want to highlight that it is subject to change. 

### Add lifecycle badge (experimental)

```{r}
usethis::use_lifecycle_badge("experimental")
```


## Phase 2: Listen & iterate 🚧 

Try to get feedback from experts on API, implementation, default decisions.  Is there already work that solves this problem? 


## Phase 3: Let things settle

### Settle on examples.  Put them in the roxygen skeleton and readme. 🚧 

### Written formal tests of functions? 🚧 


That would look like this...

```{r test_calc_frequency_works, eval = F}
library(testthat)

test_that("calc frequency works", {
  expect_equal(calc_frequency("A", 0), 440)
  expect_equal(calc_frequency("A", -1), 220)
  
})
```


```{r, eval=F}
readme2pkg::chunk_to_tests_testthat("test_calc_frequency_works")
```


### Have you worked added a description and author information in the DESCRIPTION file? 🚧 

### Addressed *all* notes, warnings and errors. 🚧 

## Promote to wider audience...

### Package website built? 🚧 

### Package website deployed? 🚧 

## Phase 3: Harden/commit

### Submit to CRAN? Or don't.  🚧 

# Appendix: Reports, Environment

## Description file extract

```{r}

```

## Environment

Here I just want to print the packages and the versions

```{r}
all <- sessionInfo() |> print() |> capture.output()
all[11:17]
```

## `devtools::check()` report

```{r, error = T, results="hide", warning=F}
devtools::check(pkg = ".")
```




